---
title: "drought_analysis_v2"
output: html_document
---

Loading libraries

```{r}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
library(skimr)
library(ggthemes)
library(naniar)
library(readr)
library(readxl)
```

## Height and Size Analysis: Pre-treatment and during treatment

Loading data:

```{r}

pre_trt_height = readxl::read_xlsx("data/cbs_drought_project.xlsx", 
                           sheet = "Location & Height Pre-Treatment") %>% 
  select(-`Number Leaves`, -`Longest Leaf (cm)`) %>% 
  rename(Height_pretrt_cm = `Height (cm)`, pop = Population, rep = Replicate, notes_predrought = notes) %>% 
  rename_all(.funs = tolower) %>% 
  # Let's just get rid of imposters here. 
  filter(is.na(notes_predrought)) %>% 
  select(-notes_predrought)

pre_trt_height

```
Pre treatment measurements were made on 9/30/2020. Number of leaves should have been completed, but longest leaf measurements were most likely forgone for lack of follow-up measurements (not enough time/energy to continue follow-up).
MB: When you have time, why don't you see if the number of leaves from 9/30 is in the datasheets. No problem if not! There was A LOT to measure. 
Blocks and racks in the pretrt height are their "final" positions after the rearrangement to make a full six racks.


```{r}

weekly_size = readxl::read_xlsx("data/cbs_drought_project.xlsx", sheet = "Weekly Surveys (complete racks)") %>% 
  filter(!is.na(rep)) %>% 
  # Let's just get rid of imposters here. 
  filter(is.na(notes)) %>% 
  select(-notes)

summary(weekly_size)

```

```{r}

weekly_size_tall_sorted <- weekly_size %>% 
  left_join(pre_trt_height) %>% 
  select(-position, -block, -rack) %>% 
  rename(`09_30_height_cm` = height_pretrt_cm) %>% 
  pivot_longer(`10_09_height_cm`:`09_30_height_cm`) %>% 
  separate(name, 6, into = c("date", "variable")) %>% 
  pivot_wider(id_cols = c(pop, age, rep, treatment, date), names_from = variable, values_from = value, values_fn = mean) %>% 
  # Let's not use the function below, because it turns numeric columns into characters
  # mutate_all(tolower) %>% 
  mutate(date = as.Date(paste0(date, 2020), format = "%m_%d_%Y"),
         age = case_when(age == "A1" ~ "7wk",
                         age == "A2" ~ "5wk",
                         age == "A3" ~ "3wk"),
         elevation = case_when(pop == "SQ1" | pop == "WL2" | pop == "YO1" | pop == "SQ3" ~ "High",
                               pop == "DPR" | pop == "KC2" | pop == "WL1" ~ "Intermediate",
                               pop == "TM2" | pop == "BH" | pop == "IH" ~ "Low"),
         # Here is a day of experiment column that is consistent with the wilting dates frame.
         # But see note later, I was not sure still quite what day 0 should be.
         day_of_experiment = as.numeric(date - as.Date("09_30_2020", format = "%m_%d_%Y")),
         height_cm = as.numeric(height_cm),
         numleaf = as.numeric(numleaf))

weekly_size_tall_sorted



```

MB: I combined the chunk below with the one above. The issue, as you discovered, was that notes was interspersed with the size columns, but since the only note was about imposters, I just got rid of those in the data loading steps and dropped the notes columns up there. 

```{r}
# 
# weekly_size_tall <- weekly_size %>% 
#   left_join(pre_trt_height) %>% 
#   select(-position, -block, -rack)
# 
# col_order <- c("pop", "age", "rep", "treatment", "10_09_height_cm", "10_09_numleaf", "10_18_height_cm",  "10_18_numleaf", "10_23_height_cm", "10_23_numleaf", "10_30_height_cm", "10_30_numleaf", "height_pretrt_cm", "notes")
# 
# weekly_size_tall <- weekly_size_tall[, col_order]
# 
# weekly_size_tall_sorted <- weekly_size_tall %>% 
#   rename(`09_30_height_cm` = height_pretrt_cm) %>% 
#   pivot_longer(`10_09_height_cm`:`09_30_height_cm`) %>% 
#   separate(name, 6, into = c("date", "variable")) %>% 
#   pivot_wider(id_cols = c(pop, age, rep, treatment, date), names_from = variable, values_from = value, values_fn = mean) %>% 
#   mutate(date = as.Date(paste0(date, 2020), format = "%m_%d_%Y")) %>%
#   mutate_all(tolower) %>% 
#   mutate(age = case_when(age == "a1" ~ "7wk",
#                          age == "a2" ~ "5wk",
#                          age == "a3" ~ "3wk")) %>%
#    mutate(elevation = case_when(pop == "sq1" | pop == "wl2" | pop == "yo1" | pop == "sq3" ~ "high",
#                                pop == "dpr" | pop == "kc2" | pop == "wl1" ~ "intermediate",
#                                pop == "tm2" | pop == "bh" | pop == "ih" ~ "low"))
# 
# weekly_size_tall_sorted  

```


```{r}

wilting_dates = read_xlsx("data/cbs_drought_project.xlsx", 
                           sheet = "Weekly Surveys (Flipped)") %>% 
  rename(pop = Population, rep = Replicate) %>% 
  rename_all(.funs = tolower)

table(wilting_dates$notes)

wilting_dates_clean = wilting_dates %>% 
  filter(is.na(notes))

```

The numbers in this data sheet correspond to the day of the experiment: `day 0`, `day 2`, and so on and so forth. On a calendar, the first day of the experiment was 10/01/2020 and the first day of observations was 10/09/2020.
MB: there are a lot of values <9 in the data. So is day 0 (or day 1?) October 1 or October 9?
Answer: Day 0 = Oct 9


Exploring data:

```{r}
summary(pre_trt_height)
```


*One height measurement is missing*: Yes, KC2-A1-10 is missing. I will look into my data sheets to see if it was recorded.

```{r}
summary(weekly_size)
```

*Let's make sure to write down metadata about how each of these was scored, i.e., what does a leaning vs. curled vs. wrinkled vs. wilted plant look like?* Great idea: here's a rough list so I don't forget later.
+ `Leaning` phenotypes were designated when the upper half of the plant (closer to the top) were beginning to lean down at ~45-65 degree angles (were not "standing up straight").
+ `Curled` phenotypes were designated when the plant leaves began to curl in on themselves.
+ `Wrinkled` phenotypes were designated when the plant leaves began to show signs of wrinkling (e.g. no longer smooth or flat).
+ `Wilted` phenotypes were designated when the plant was "dry" and crisp to the touch.
+ `Color` phenotypes were designated based on the color of the plant at the time of observation. Dates were written when the plant was beginning to show a *new* phenotype only, i.e., it was recorded once that a plant transitioned to redgreen and once recorded when transitioned to red.

*Also, let's document what the numbers mean in the weekly surveys. Is it day of month or day of expt. and if day of expt, what is day 1?*

```{r}
table(pre_trt_height$pop)
table(pre_trt_height$treatment)
```

```{r}
boxplot(pre_trt_height$height_pretrt_cm ~ pre_trt_height$pop)
```


```{r}
boxplot(weekly_size$`10_09_height_cm` ~ weekly_size$pop)
```

```{r}
hist(weekly_size$`10_18_height_cm`, breaks = 40)
```
```{r}
boxplot(weekly_size$`10_18_height_cm` ~ weekly_size$pop)
```

```{r}
hist(weekly_size$`10_23_height_cm`, breaks = 40)
```

```{r}
boxplot(weekly_size$`10_23_height_cm` ~ weekly_size$pop)
```

```{r}
hist(weekly_size$`10_30_height_cm`, breaks = 40) 
```

```{r}
boxplot(weekly_size$`10_30_height_cm` ~ weekly_size$pop)
```

<!-- Resolved data entry issues -->
<!-- *WL2 is over-represented by 1, WL1 is underrepresented by 1, is this a mix-up?* -->
<!-- *Occurs in block 10, Age 3...* It has been corrected as of 4/4/2021: The mix-up was due to a typo. -->
<!-- *There is also an instance of 2 IH A1 seedlings in block 12 and no IH A3 seedling in block 12* It has been corrected as of 4/4/2021: The mix-up was due to a typo on the label. -->
<!-- *Are blocks/racks in pretrt height the final positions, or positions before rearrangement?* Blocks and racks in the pretrt height are their "final" positions after the rearrangement to make a full six racks. -->
<!-- *Outlier at 29. Data entry error or one of the non-streptanthus tricksters?* Fixed as of 4/4/2021: data entry error (transposed a `.`) -->
<!-- *Outlier at 21.5. Data entry error or one of the non-streptanthus tricksters?* Indeed, a non-Steptanthus trickster! SQ3-A2-10. -->
<!-- *When weekly size is made tall, get an error message about non-unique values. Why?* Fixed on 4/4/2021. See above. -->
<!-- *We should finish coding these consistently-some are numbers and some are dates with an underscore still.* Fixed on 4/4/2021. -->
<!-- *One age is "1".* Fixed on 4/4/2021. Data entry error. -->


```{r}
summary(weekly_size_tall_sorted)
```






## Weekly Survey Analysis: Phenotypic Differences

Loading data:

```{r}

weekly_pheno <- readxl::read_xlsx("data/cbs_drought_project.xlsx", sheet = "Weekly Surveys (Flipped)")
weekly_pheno
table(weekly_pheno$notes)

```

Numbers in the phenotype columns denote the numeric (and not calendar day) in which the phenotype was observed (e.g. day 0, day 17, etc.)

Cleaning data

```{r}

weekly_pheno_cleaned <- weekly_pheno %>% 
  # Get rid of imposters
  filter(is.na(notes)) %>% 
  clean_names() %>% 
  # Make consistent with other dataframe
  rename(pop = population, rep = replicate) %>% 
  # Let's not use the function below, because it turns numeric columns into characters
  # mutate_all(tolower) %>% 
  mutate(age = case_when(age == "A1" ~ "7wk",
                         age == "A2" ~ "5wk",
                         age == "A3" ~ "3wk"),
         elevation = case_when(pop == "SQ1" | pop == "WL2" | pop == "YO1" | pop == "SQ3" ~ "High",
                               pop == "DPR" | pop == "KC2" | pop == "WL1" ~ "Intermediate",
                               pop == "TM2" | pop == "BH" | pop == "IH" ~ "Low"),
         age = as.factor(age),
         pop = as.factor(pop)) %>% 
  select(-rack, -block, -position) %>% 
  # Going to go ahead and get rid of unassigned plants
  filter(pop != "NONE")

weekly_pheno_cleaned

```

Checking data

```{r}

tabyl(weekly_pheno_cleaned, pop)

```
There are 10 populations present, with `none` corresponding to replicates of a population removed from their original position in the rack (due to either a lack of growth, or shuffling to complete priority racks).

```{r}

summary(weekly_pheno_cleaned)

```

```{r}
table(weekly_pheno_cleaned$pop)

# Which phenotypes occur frequently in each treatment?

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$green))
# Not really

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$yellowgreen))
# Frequent in severe drought

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$redgreen))

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$yellow))

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$brown))

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$curled))

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$wrinkled))

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$leaning))

table(weekly_pheno_cleaned$treatment, is.na(weekly_pheno_cleaned$wilted))

```

Make plots!

We could consider making some composite columns, ie. the first day any discoloration was observed. I can help with this, it would require making data tall, grouping by plant, then taking the minimum value in each group. 

First, just doing the wilted response. 
In this case, I think it's only worth plotting the severe drought treatment, because there was only one instance of wilting in the other treatments. 

```{r}
  
# The simplest option is to use the ECDF option in ggplot. 
# But, a trick: it removes NAs, so by default would drop plants that never wilted. We don't want that!
# Can add a dummy value of a date after all others, but then I would crop this last value out when presenting

weekly_pheno_cleaned %>% 
  filter(treatment == "Severe drought",
         elevation == "Low") %>% 
  mutate(wilted = ifelse(is.na(wilted), 25, wilted)) %>%
  ggplot() +
  stat_ecdf(aes(x = wilted, color = age), geom = "step") +
  facet_wrap(.~pop, nrow = 1)
ggsave("plots/wilting_over_time_low.pdf", height = 3, width = 8)


weekly_pheno_cleaned %>% 
  filter(treatment == "Severe drought",
         elevation == "Intermediate") %>% 
  mutate(wilted = ifelse(is.na(wilted), 25, wilted)) %>%
  ggplot() +
  stat_ecdf(aes(x = wilted, color = age), geom = "step") +
  facet_wrap(.~pop, nrow = 1)
ggsave("plots/wilting_over_time_int.pdf", height = 3, width = 8)


weekly_pheno_cleaned %>% 
  filter(treatment == "Severe drought",
         elevation == "High") %>% 
  mutate(wilted = ifelse(is.na(wilted), 25, wilted)) %>%
  ggplot() +
  stat_ecdf(aes(x = wilted, color = age), geom = "step") +
  facet_wrap(.~pop, nrow = 1)
ggsave("plots/wilting_over_time_high.pdf", height = 3, width = 11)



# Alternative approach: use cumsum
# This is a bit complicated but gives a bit more plotting flexibility

dummy = crossing(pop = unique(weekly_pheno_cleaned$pop), 
                 # treatment = unique(weekly_pheno_cleaned$treatment), 
                 age = unique(weekly_pheno_cleaned$age),
                 wilted = unique(c(weekly_pheno_cleaned$wilted, 0))) %>% 
  filter(!is.na(wilted))

weekly_pheno_cleaned %>% 
  filter(treatment == "Severe drought") %>% 
  arrange(pop, age, treatment, wilted) %>% 
  mutate(wiltedyn = !is.na(wilted)) %>% 
  group_by(pop, age, treatment, wilted) %>% 
  summarize(num_wilted = sum(wiltedyn), n = n()) %>% 
  group_by(pop, age, treatment) %>% 
  mutate(cumulative_wilted = cumsum(num_wilted), 
         n = sum(n),
         prop_wilted = cumulative_wilted/n) %>% 
  ggplot() +
  geom_point(aes(x = wilted, y = prop_wilted, color = age)) +
  geom_line(aes(x = wilted, y = prop_wilted, color = age)) +
  facet_wrap(.~pop)

# Need to pull in rack data eventually as a random effect in models.

```

Plot final proportions wilting

```{r}
# check =
weekly_pheno_cleaned %>% 
  group_by(pop, age, treatment) %>% 
  summarize(total = n(), count_wilted = sum(!is.na(wilted))) %>% 
  mutate(prop_wilted = count_wilted/total) %>% 
  filter(treatment == "Severe drought") %>% 
  ggplot() +
  geom_col(aes(x = pop, y = prop_wilted, fill = age), position = position_dodge()) +
  facet_wrap(.~treatment)


```


Plot by age, size, treatment

```{r}

weekly_size_tall_sorted %>% 
  filter(day_of_experiment == 30,
         height_cm < 20) %>% 
  ggplot() +
  geom_boxplot(aes(x = age, fill = treatment, y = height_cm)) +
  facet_wrap(.~pop, scales = "free_y")


weekly_size_tall_sorted %>% 
  filter(day_of_experiment == 30,
         height_cm < 20,
         elevation  == "Low") %>% 
  ggplot() +
  geom_boxplot(aes(x = treatment, fill = age, y = height_cm)) +
  facet_wrap(.~pop, scales = "free_y")
ggsave("plots/height_trt_age_low.pdf", height = 3, width = 10)

weekly_size_tall_sorted %>% 
  filter(day_of_experiment == 30,
         height_cm < 20,
         elevation  == "Intermediate") %>% 
  ggplot() +
  geom_boxplot(aes(x = treatment, fill = age, y = height_cm)) +
  facet_wrap(.~pop, scales = "free_y")
ggsave("plots/height_trt_age_int.pdf", height = 3, width = 10)

weekly_size_tall_sorted %>% 
  filter(day_of_experiment == 30,
         height_cm < 20,
         elevation  == "High") %>% 
  ggplot() +
  geom_boxplot(aes(x = treatment, fill = age, y = height_cm)) +
  facet_wrap(.~pop, nrow = 1, scales = "free_y")
ggsave("plots/height_trt_age_high.pdf", height = 3, width = 14)

```

Leaf counts

```{r}
weekly_size_tall_sorted %>% 
  filter(day_of_experiment == 30,
         elevation  == "Low") %>% 
  group_by(pop, treatment, age) %>% 
  summarize(mean_leaf_num = mean(numleaf)) %>% 
  ggplot() +
  geom_col(aes(x = treatment, fill = age, y = mean_leaf_num), position = position_dodge()) +
  facet_wrap(.~pop, scales = "free_y")
ggsave("plots/leaves_trt_age_low.pdf", height = 3, width = 10)

weekly_size_tall_sorted %>% 
  filter(day_of_experiment == 30,
         elevation  == "Intermediate") %>% 
  group_by(pop, treatment, age) %>% 
  summarize(mean_leaf_num = mean(numleaf)) %>% 
  ggplot() +
  geom_col(aes(x = treatment, fill = age, y = mean_leaf_num), position = position_dodge()) +
  facet_wrap(.~pop, scales = "free_y")
ggsave("plots/leaves_trt_age_int.pdf", height = 3, width = 10)

weekly_size_tall_sorted %>% 
  filter(day_of_experiment == 30,
         elevation  == "High") %>% 
  group_by(pop, treatment, age) %>% 
  summarize(mean_leaf_num = mean(numleaf)) %>% 
  ggplot() +
  geom_col(aes(x = treatment, fill = age, y = mean_leaf_num), position = position_dodge()) +
  facet_wrap(.~pop, scales = "free_y", nrow = 1)
ggsave("plots/leaves_trt_age_high.pdf", height = 3, width = 13)

```



