---
title: "regressions_practice"
output: html_document
---

#Swirl Course

```{r}
swirl::install_course("Regression Models")
```

```{r}
swirl()
```


#UC Davis Linear Models

##What is a linear model?

A linear model is a model for a continuous outcome `y` of the form: 
`Y=β0+β1X1+β2X2+⋯+βpXp+ϵ`

The covariates `x` can be a continuous variable (age, weight, temperature) or a dummy variable coding for a categorical covariate.

`ϵ` is assumed to be normally distributed with constant variance across range of data.

Models with *all categorical covariates* are ANOVA models. Models with *continuous covariates* are linear regression models. They're both linear models, so R does not distinguish between them.

```{r}
linear_model_ex <- read.csv("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2018-September-Bioinformatics-Prerequisites/master/friday/lm_example_data.csv")

head(linear_model_ex)
```

```{r}
str(linear_model_ex)
```

```{r}
summary(linear_model_ex)
```

Now that we have data, fit it to a model using `expression` as the outcome and `treatment` as a categorical covariate.
```{r}
oneway_model <- lm(expression ~ treatment, data = linear_model_ex)
```

```{r}
oneway_model
```

The outcome of the model is on the left side, with covariates separated by `+` following the `~` symbol.

```{r}
class(oneway_model)
```

Using `summary` on an `lm` object will give p-values and other relevant information.
```{r}
summary(oneway_model)
```

What does it all mean?
+ Coefficients are the `β`
+ Estimate is the estimate of each coefficient
+ Std. error is the standard error of the estimate
+ `t value` is the coefficient divided by its standard error
+ Pr(>|t|) is the p-value for the coefficient
+ The residual standard error is estimate of variance of `ϵ`
+ Degrees of freedom is the sample size minus the number of coefficients estimated
+ R-squared is (roughly) the proportion of variance in the outcome explained by the model
+ The F-statistic compares the fit of the model *as a whole* to the null model (with no covariates)

`coef()` gives you model coefficients:
```{r}
coef(oneway_model)
```

What do `model coefficients` mean?
R uses reference group coding or 'treatment contrasts'. For categorical covariates, the first level alphabetically (or first level factor) is the `reference group`. The reference group does not have its own coefficient and is represented by the intercept instead. Coefficients for other groups are the *difference from the reference*:

In this example:
+ `Intercept` is the *mean of expression for treatment* = A
+ `treatmentB` is the mean expression for treatment B minus the mean for treatment A
+ `treatmentC` is the mean expression for treatment C minus the mean for treatment A

If you want to get the means in each treatment:
```{r}
treatmentmeans <- tapply(linear_model_ex$expression, linear_model_ex$treatment, mean)
treatmentmeans["A"]
```

If you want to find the difference in means for your coefficient:
```{r}
treatmentmeans["B"] - treatmentmeans["A"]
```

If you do not want to reference group coding, you can fit the model without an intercept:
```{r}
no_intercept_model <- lm(expression ~ 0 + treatment, data = linear_model_ex) # 0 = no intercept
summary(no_intercept_model)
```

