---
title: "Model fitting"
author: "Megan Bontrager"
date: "3/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mosaic)
library(readxl)
library(lme4)
library(lmerTest)
library(ggeffects)

```

## Load data

```{r data}

pre_trt_height = read_xlsx("data/cbs_drought_project.xlsx", 
                           sheet = "Location & Height Pre-Treatment") %>% 
  select(-`Number Leaves`, -`Longest Leaf (cm)`) %>% 
  rename(Height_pretrt_cm = `Height (cm)`, pop = Population, rep = Replicate) %>% 
  rename_all(.funs = tolower)

weekly_size = read_xlsx("data/cbs_drought_project.xlsx", 
                        sheet = "Weekly Surveys (complete racks)") %>% 
  filter(!is.na(rep))

# Make tall
weekly_size_tall = weekly_size %>% 
  left_join(pre_trt_height) %>% 
  select(-position, -block, -rack) %>% 
  # NEED TO UPDATE BELOW DATE WITH ACTUAL DATE OF PRE-TRT MEASUREMENTS
  rename(`09_20_height_cm` = height_pretrt_cm) %>% 
  pivot_longer(`10_09_height_cm`:`09_20_height_cm`) %>% 
  separate(name, 6, into = c("date", "variable")) %>% 
  pivot_wider(id_cols = c(pop, age, rep, treatment, date), names_from = variable, values_from = value, values_fn = mean) %>% 
  mutate(date = as.Date(paste0(date, 2020), format = "%m_%d_%Y"))
# May want to make a day of expt column or something to match wilting dates

wilting_dates = read_xlsx("data/cbs_drought_project.xlsx", 
                           sheet = "Weekly Surveys (Flipped)") %>% 
  rename(pop = Population, rep = Replicate) %>% 
  rename_all(.funs = tolower) #%>% 
  # mutate(wilted = as.numeric(wilted))
# Some of these are still coded as month_day, let's make them consistent


```

## Explore data, check for errors

Wondering if we have:
- Takedown data

```{r data validation, eval = FALSE}

summary(pre_trt_height)
# What date were these measurements made on and were the # leaves and longest leaf measurements completed?
# One height measurement is missing
table(pre_trt_height$pop)
table(pre_trt_height$treatment)
# WL2 is over-represented by 1, WL1 is underrepresented by 1, is this a mix-up?
table(pre_trt_height$pop,pre_trt_height$block, pre_trt_height$age)
# Occurs in block 10, Age 3
# There is also an instance of 2 IH A1 seedlings in block 12 and no IH A3 seedling in block 12
summary(pre_trt_height)
boxplot(pre_trt_height$height_pretrt_cm ~ pre_trt_height$population)
# Are blocks/racks in pretrt height the final positions, or positions before rearrangement?

summary(weekly_size)
hist(weekly_size$`10_09_height_cm`, breaks = 40)
boxplot(weekly_size$`10_09_height_cm` ~ weekly_size$pop)
# Outlier at 29. Data entry error or one of the non-streptanthus tricksters?
hist(weekly_size$`10_18_height_cm`, breaks = 40) # Looks good
boxplot(weekly_size$`10_18_height_cm` ~ weekly_size$pop)
hist(weekly_size$`10_23_height_cm`, breaks = 40) # Looks good
boxplot(weekly_size$`10_23_height_cm` ~ weekly_size$pop)
hist(weekly_size$`10_30_height_cm`, breaks = 40) 
boxplot(weekly_size$`10_30_height_cm` ~ weekly_size$pop)
# Outlier at 21.5. Data entry error or one of the non-streptanthus tricksters?

# When weekly size is made tall, get an error message about non-unique values. Why?
check = weekly_size %>% 
  group_by(pop, age, rep, treatment) %>% 
  summarize(n = n())
# It's because these two identities are duplicated:
# IH	A1	12	Light drought	
# WL2	A3	10	Control	

summary(wilting_dates)
# We should finish coding these consistently-some are numbers and some are dates with an underscore still. 
# Let's make sure to write down metadata about how each of these was scored, i.e., what does a leaning vs. curled vs. wrinkled vs. wilted plant look like?
# Also, let's document what the number means. Is it day of month or day of expt. and if day of expt, what is day 1?
table(wilting_dates$age)
table(wilting_dates$wilted)
# One age is "1"

```



## Q1: How do seedlings of different sizes respond to moderate and severe drought?

Response variables: 
- change in size over time
- final size
- wilting date (in severe trt)

```{r}

final_size = weekly_size_tall %>% 
  filter(date == "2020-10-30") %>% 
  filter(height_cm > 0)

hist(final_size$height_cm, breaks = 20)
hist(log(final_size$height_cm))

mod_fs = lmer(log(height_cm) ~ age*treatment + (1|pop), data = final_size)
summary(mod_fs)

plot_fs = ggeffect(mod_fs, terms = c("age", "treatment"))

plot(plot_fs)


# Need to set this up as time-to-event model
wilting_dates_severe = wilting_dates %>% 
  filter(treatment == "Severe drought", age != 1) %>% 
  mutate(wilted = ifelse(is.na(wilted), 30, wilted))

summary(wilting_dates_severe)

mod_wd = lmer(wilted ~ age + (1|pop), data = wilting_dates_severe)
summary(mod_wd)

plot_wd = ggeffect(mod_wd, terms = c("age"))

plot(plot_wd)

```




## Q2: Are there differences among populations in seedlings' ability to withstand drought?



## Q3: Can these responses (and their implications) be better understood in the context of weather patterns from sites-of-origin?
